---
title: "NAWQA Nutrients: Data Preparation for Land Use vs N & P Course Project"
output: html_document
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, message = TRUE, warning = TRUE)
```

## Introduction

This short report documents the preparation of a small, tidy dataset derived from the NAWQA nutrients repository (Manning et al., 2020). The objective is to obtain a compact, reproducible subset suitable for a course project on the impact of watershed land-use types on downstream nitrogen (TN) and phosphorus (TP) concentrations.

We follow the original repository’s R Markdown logic to load the raw data and reproduce the principal joins that combine nutrient measurements, station coordinates, and StreamCat land-use percentages. We then aggregate to one record per station and draw a small, balanced random sample for teaching.

## Data Sources

- `NAWQA.NP` – nutrient concentrations at stream stations.
- `nawqa.latlong` – station coordinates (latitude/longitude and identifiers).
- `nawqa_landuse` – watershed land-use percentages (StreamCat, 2006).

Loading and joins mirror the patterns in the original Rmd files contained in this repository (e.g., `nawqa_EAP_rmd_2020_02_26.Rmd`).

## Variable Selection and Processing Steps

- Join nutrients, coordinates, and land-use into a single data frame (`nawqa_full`).
- Construct land-use summaries (percent of watershed):
  - `pAG = PctCrop2006Ws + PctHay2006Ws`
  - `pURB = PctUrbLo2006Ws + PctUrbMd2006Ws + PctUrbHi2006Ws`
  - `pFOR = PctConif2006Ws + PctDecid2006Ws + PctMxFst2006Ws`
- Define `LandUseType` using simple thresholds (dominant class):
  - Agricultural if `pAG >= 50` and `pAG >= pURB` and `pAG >= pFOR`.
  - Urban if `pURB >= 20` and `pURB >= pAG` and `pURB >= pFOR`.
  - Forest if `pFOR >= 50` and `pFOR >= pAG` and `pFOR >= pURB`.
  - Otherwise, Mixed.
- Collapse multiple samples per station to a single record using station-level medians of TN and TP (NO3 and NH4 included if available).
- Filter to stations with positive, non-missing TN and TP and with `LandUseType` ∈ {Agricultural, Urban, Forest}.
- Draw a balanced random sample (up to 40 stations per land-use type).
- Save the final dataset to `data/nawqa_landuse_nutrients_subset.csv`.

## Load Prepared Data

```{r load-prepared}
suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
})

csv_path <- "data/nawqa_landuse_nutrients_subset.csv"
if (!file.exists(csv_path)) {
  message("Prepared CSV not found — sourcing the preparation script...")
  source("prepare_nawqa_subset.R")
}

nawqa_course_data <- readr::read_csv(csv_path, show_col_types = FALSE)

summary_by_landuse <- nawqa_course_data %>%
  group_by(LandUseType) %>%
  summarise(
    n = dplyr::n(),
    mean_TN = mean(TN, na.rm = TRUE),
    sd_TN   = sd(TN, na.rm = TRUE),
    mean_TP = mean(TP, na.rm = TRUE),
    sd_TP   = sd(TP, na.rm = TRUE),
    .groups = "drop"
  )
```

## Descriptive Statistics

```{r describe}
if (requireNamespace("knitr", quietly = TRUE)) {
  knitr::kable(summary_by_landuse, caption = "Summary by Land Use Type")
} else {
  print(summary_by_landuse)
}
```

Optional: a simple visualization (TN by land-use type).

```{r optional-plot, fig.width=6, fig.height=4}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)
  ggplot(nawqa_course_data, aes(x = LandUseType, y = TN)) +
    geom_boxplot(outlier.alpha = 0.4) +
    labs(x = "Land Use Type", y = "TN (mg/L)",
         title = "TN by Land Use Type (sampled stations)") +
    theme_minimal()
}
```

## Output Dataset Description

- File: `data/nawqa_landuse_nutrients_subset.csv`
- Dimensions: the file contains `r nrow(nawqa_course_data)` rows and `r ncol(nawqa_course_data)` columns.
- Columns:
  - `stationId` – station identifier.
  - `statePostalCode` – two-letter state code.
  - `placeName` – station location name.
  - `latitude`, `longitude` – station coordinates.
  - `LandUseType` – categorical class (Agricultural, Urban, Forest).
  - `pAG`, `pURB`, `pFOR` – summarized land-use percentages (agriculture, urban, forest).
  - `TN`, `TP` – median total nitrogen and total phosphorus (mg/L) per station.
  - `NO3`, `NH4` – median nitrate+nitrite and ammonium (mg/L) per station, if available.

